Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-17 paper/paper.tex
clarifying some things + integrating some comments

\newtheorem{theorem}{Theorem} \newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}{Remark}

@@

Model-free RL is plausible in the human brain. Indeed,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.

In our present model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

\begin{definition}
\textbf{XXX Danilo: Is there a definition 2 somewhere?}
Mathematically, a POMDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts off in state $s$.
\end{itemize}

% \begin{remark}
%   Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
%   except the Markov property, we can always transform it into an MDP by considering a compounded
%   version of the states
%   $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
%   history up to an including time $t$.
% \end{remark}
\end{definition}

@@

Computing an optimal policy then corresponds to finding the parameters $\theta_{k}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal D}[(\tilde{Q}(s, a|\theta) - \tilde{Q}(s',a'|\theta^Q_{k}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal D}\left[\frac{1}{2}(\tilde{Q}(s, a|\theta_{k}) - y_k)^2\right].
  \label{eq:oracle}
\end{equation}
\textbf{We still may need to explain y in the formula with one sentence.}
For instance, a general linear model with a kernel $\phi$ would be of the
form
$$\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
A recently proposed, practically successful alternative approach
\citep{mnih2015,silver2016mastering} is to learn this
representation using a deep neural network, leading to the so-called \textit{Deep Q-learning} family of methods which
likely are state-of-the-art in RL.
The set of $\theta$ parameters that instantiate the nonlinear interactions
between layers of the artifical neural network
may find a neurobiological correspondence in the adaptive strengths of axonal
connections between neurons from the different levels
of the neural processing hierarchy
\citep{mesulam1998sensation, taylor2015global}.

@@

\section{Relation to other models decision-making and optimal control for the brain}
\subsection{The free-energy principle and active inference \
textbf{move to Appendix}}
% \subsubsection{The theory}
In Friston's free-energy principle (FEP) \citep{friston2010free,fristonAIorRL}, the brain is portrayed
as biomechanical inference engine which must minimize the long-term average of surprise.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net (DBN)}. According to FEP's account, the goal of the brain is then to optimize over the generative model $G$: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics, etc., so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles. These is called the \textit{generative} model. FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams / fantasizes imaginary worlds $\z$ whilst the latter tries to generate sensations $\x$ which match these dreams!

@@

\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-modal and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
from stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of sensory sensation is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal back projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, generative nature of sensory perception
\citep{friston2010free} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual sensorimotor integration of perception-action cycles
may be maintained by top-down modulation
by a-priori information about the environment.

>>>>

\newtheorem{theorem}{Theorem} \newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{definition*}{Definition}
\newtheorem{remark}{Remark}

@@

Model-free RL is plausible in the human brain. Indeed,
it has long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
In our present model-free RL framework,
a rat might represent such knowledge about the world as follows:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

\begin{mdframed}
\begin{definition*}
% \textbf{XXX Danilo: Is there a definition 2 somewhere?}
Mathematically, a POMDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts off in state $s$.
\end{itemize}

% \begin{remark}
%   Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
%   except the Markov property, we can always transform it into an MDP by considering a compounded
%   version of the states
%   $$S_t \leftarrow (S_0,A_0,R_0,S_1,A_1,S_2\ldots,S_t,A_{t-1},R_{t-1},S_{t})$$ made of the system's
%   history up to an including time $t$.
% \end{remark}
\end{definition*}
\end{mdframed}

@@

Computing an optimal policy then corresponds to finding the parameters $\theta_{k}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal D}[(\tilde{Q}(s, a|\theta) - \tilde{Q}(s',a'|\theta^Q_{k}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal D}\left[\frac{1}{2}(\tilde{Q}(s, a|\theta_{k}) - y_k)^2\right],
  \label{eq:oracle}
\end{equation}
where $y_k$ is a defined in \eqref{eq:bellman}
% \textbf{We still may need to explain y in the formula with one sentence.}
For instance, a general linear model with a kernel $\phi$ would be of the
form
$$\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
A recently proposed, practically successful alternative approach
\citep{mnih2015,silver2016mastering} is to learn this
representation using a deep neural network, leading to the so-called \textit{Deep Q-learning} family of methods which
likely are state-of-the-art in RL.
The set of $\theta$ parameters that instantiate the nonlinear interactions
between layers of the artifical neural network
may find a neurobiological correspondence in the adaptive strengths of axonal
connections between neurons from the different levels
of the neural processing hierarchy
\citep{mesulam1998sensation, taylor2015global}.

@@

\section{Relation to other models decision-making and optimal control for the brain}
\subsection{Predictive coding hypothesis}
Predictive coding mechanisms
\citep{clark2013whatever, friston2008hierarchical}
are a frequently evoked idea in the context of default mode function
\citep{bar2007}.
Cortical responses are explained as
emerging from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
Feed-forward sensory processing is constantly calibrated by
top-down modulation from more multi-modal and associative brain regions
further away from primary sensory cortical regions.
The dynamic interplay between cortical processing levels
may enable inferences about the world by reconciling
gaps between fresh sensory input and expectations computed
from stored prior information.
At each stage of neural processing,
an internally generated prediction of aspects of sensory sensation is
directly compared against the actual environmental input.
A prediction error at one of the processing levels
incurs plasticity changes of neuronal back projections
(i.e., adapting model parameters)
to allow for gradually improved future prediction of the environment.
In this way,
the predictive coding hypothesis offers explanations for
the constructive, generative nature of sensory perception
\citep{friston2010free} and
the intimate relation of motor action to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
Contextual sensorimotor integration of perception-action cycles
may be maintained by top-down modulation
by a-priori information about the environment.

\subsection{The free-energy principle and active inference \textbf{XXX move to Appendix}}
% \subsubsection{The theory}
In Friston's free-energy principle (FEP) \citep{friston2010free,fristonAIorRL}, the brain is portrayed
as biomechanical inference engine which must minimize the long-term average of surprise.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net (DBN)}. According to FEP's account, the goal of the brain is then to optimize over the generative model $G$: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics, etc., so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles. These is called the \textit{generative} model. FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams / fantasizes imaginary worlds $\z$ whilst the latter tries to generate sensations $\x$ which match these dreams!

@@
